---
title: "Structural Break and Time Series Modeling of Weekly Cereal Sales"
author: "Alex Spigler"
date: "April 4, 2025"
output:
  pdf_document:
    toc: false
    number_sections: false
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 6.5,
  fig.height = 4
)

library(astsa)     # time series tools: acf2(), sarima(), sarima.for()
library(lmtest)    # Durbin–Watson and RESET tests
library(car)       # ncvTest()
library(tseries)   # kpss.test()
```

# Introduction

This report analyzes weekly cereal sales for a brand over a two-year period. At week 88, a competitor introduced a similar product, and the goal is to test whether that entry shifted the level of sales, the trend, or both. The data is synthetic, generated to demonstrate the intervention-analysis workflow; the focus is the modeling and diagnostics rather than the specific series.

The goals of this analysis are to:

1.  Model the underlying trend and a possible level/trend shift at the intervention week.
2.  Diagnose any violations of classical regression assumptions, especially autocorrelation.
3.  Produce short-term forecasts with prediction intervals.

The analysis is carried out in R using a regression model with an intervention term and ARIMA errors.

# Data and Exploratory Analysis

```{r load-data}
cereal <- read.csv("cereal-sales-data.csv")
cereal_ts <- ts(cereal$sales, start = 1, frequency = 52)

# Plot with week numbers on x-axis
plot(
  x = 1:length(cereal_ts),
  y = as.numeric(cereal_ts),
  type = "l",
  main = "Weekly Cereal Sales",
  xlab = "Week",
  ylab = "Sales"
)
```

Visually, the series shows a non-constant mean and an upward trend, indicating nonstationarity. There also appears to be a noticeable change around week 88, motivating an intervention model.

This nonstationary trend could be deterministic (the series fluctuating around a fixed trend line, removed by including a time term in a regression) or stochastic (a unit root, where shocks accumulate permanently, removed by differencing). The two call for different models, so the distinction matters. Because the series also contains a structural break, the raw series cannot be tested directly: standard unit-root tests such as ADF and KPSS are unreliable when a break is present, since a break makes a deterministic trend look like a unit root. We therefore model the deterministic structure (trend plus intervention) with regression first, then check the residuals for any leftover stochastic trend that would require differencing.

# Methods

## Regression with Intervention

An intervention regression model is appropriate for modeling a structural break at a known intervention point. Let:

-   $y_t$ = cereal sales in week $t$
-   $x_t = 0$ for $t < 88$, $x_t = 1$ for $t \ge 88$
-   $t$ = week index
-   $x_t t$ = interaction term

The model is:

$$y_t = \beta_0 + \beta_1 x_t + \beta_2 t + \beta_3 x_t t + \varepsilon_t$$

This specification is appropriate because:

- $\beta_0$ captures the intercept (baseline level of sales) before week 88
- $\beta_2 t$ captures the trend (rate of change in sales) before week 88
- $\beta_1 x_t$ allows for a jump or drop in level at week 88
- $\beta_3 x_t t$ allows for a change in trend after week 88 (sales could start declining faster or growing more slowly)

```{r define-intervention}
n <- nrow(cereal)
cereal$t  <- seq_len(n)
cereal$x  <- ifelse(cereal$t < 88, 0, 1)
cereal$xt <- cereal$x * cereal$t

reg_model <- lm(sales ~ x + t + xt, data = cereal)
summary(reg_model)
```

**Interpretation of coefficients:**

All four coefficients are statistically significant (p < 0.001). Because the model includes the trend-interaction term $x_t t$, the step and interaction coefficients are not interpretable on their own; the immediate jump and the post-intervention slope each combine two coefficients.

-   $\hat{\beta}_0 = 102,585$: Baseline sales level before week 88
-   $\hat{\beta}_2 = 1,153$: Pre-intervention trend (sales increased by approximately 1,153 units per week)
-   $\hat{\beta}_1 = 278,483$: Coefficient on the step dummy $x_t$ (the gap between the two fitted lines extrapolated back to $t = 0$), not the jump at week 88
-   $\hat{\beta}_3 = -3,257$: Coefficient on the interaction $x_t t$, i.e. the change in slope at week 88, not the post-intervention slope itself

Combining the relevant terms gives the quantities of interest:

-   **Immediate level shift at week 88** $= \hat{\beta}_1 + \hat{\beta}_3 \cdot 88 = 278,483 - 286,616 \approx -8,100$ units. The intervention produces little immediate jump; the raw series is essentially flat across the break (weeks 87 to 89: 202,370, 201,100, 203,210).
-   **Post-intervention trend** $= \hat{\beta}_2 + \hat{\beta}_3 = 1,153 - 3,257 = -2,104$ units/week. Competitor entry reverses the trend from $+1,153$ units/week (growth) to $-2,104$ units/week (decline), a change in slope of $-3,257$ units/week.

# Regression Diagnostics

```{r regression-diagnostics, fig.height=8}
reg_resid <- resid(reg_model)
par(mfrow = c(2, 1))

plot(reg_resid, type="l",
     main="Residuals vs Time", xlab="Week", ylab="Residuals")
abline(h=0, col="red", lty=2)

plot(fitted(reg_model), reg_resid,
     main="Residuals vs Fitted", xlab="Fitted", ylab="Residuals")
abline(h=0, col="red", lty=2)

qqnorm(reg_resid, main="Normal Q–Q Plot")
qqline(reg_resid, col="red")

acf(reg_resid, main="ACF of Regression Residuals")

par(mfrow = c(1, 1))
```

## Tests

```{r assumption-tests}
resettest(reg_model)
dwtest(reg_model)
ncvTest(reg_model)
shapiro.test(reg_resid)
```

**Linearity:** The Ramsey RESET test (p-value = 0.5211) shows no evidence of omitted nonlinearity in the fitted trend, consistent with the residuals vs. fitted plot, which shows no systematic curvature.

**Independence:** While the residuals vs. time plot does not show an obvious visual pattern, formal testing reveals positive autocorrelation. The Durbin-Watson test (p-value = 0.001955) rejects the null hypothesis of no autocorrelation. This is further confirmed by the ACF plot, which shows a significant spike at lag 1.

**Constant Variance:** The residuals vs. fitted values plot shows residuals randomly distributed around y = 0, suggesting constant variance. This is supported by the Cook-Weisberg non-constant variance score test (p-value = 0.2911), which does not reject constant variance. The test examines whether residual variance increases with the fitted level, the form of heteroscedasticity most plausible for a trending sales series.

**Normality:** The Q-Q plot appears approximately normal, and the Shapiro-Wilk test (p-value = 0.5139) does not reject the null hypothesis of normality.

**Conclusion:** Of the four classical regression assumptions, only independence fails: the errors show positive autocorrelation. This motivates the use of ARIMA errors.

# Model Refinement with ARIMA Errors

## Residual Analysis

Because the diagnostic tests showed positive autocorrelation among the error terms, the residuals are not white noise and require further modeling.

```{r residual-analysis}
# level setting: the regression already removed the trend
kpss.test(reg_resid, null = "Level")
acf2(reg_resid, max.lag = 40)
```

**Stationarity Check:** The level setting is the correct one here: the regression has already removed the deterministic trend, so the residuals have no trend left for a trend setting to test. The KPSS test (p-value > 0.1) does not reject the null of level stationarity, consistent with residuals that have a constant mean and no remaining stochastic trend (unit root). Differencing is therefore not needed, and the leftover residual autocorrelation can be modeled directly with stationary ARMA errors.

**Seasonality Check:** The ACF shows no significant spikes at seasonal lags such as 52 or 104, indicating no seasonal autocorrelation. Seasonal differencing is not appropriate for this series.

**Model Selection:** Both the ACF and PACF show a significant spike at lag 1 and little structure beyond it, which points to a low-order model but does not sharply identify the orders. Comparing the candidate error models confirms this: AR(1), MA(1), and ARMA(1,1) all reduce the residuals to white noise (Ljung-Box p > 0.05 across the tested lags) and fall within about 2 AIC of one another. ARIMA(1,0,1) has the lowest AIC and significant AR and MA terms, so it is used here; a more parsimonious AR(1), which BIC prefers, would serve equally well and leaves the structural-break estimates and forecasts unchanged.

```{r model-comparison}
# Compare candidate low-order error models on the regression residuals
models <- list(
  "AR(1)"     = c(1, 0, 0),
  "MA(1)"     = c(0, 0, 1),
  "ARMA(1,1)" = c(1, 0, 1)
)
comparison <- data.frame(
  Model = names(models), AIC = NA, BIC = NA, LjungBox_p = NA
)

for (i in seq_along(models)) {
  ord <- models[[i]]
  fit <- arima(reg_resid, order = ord, include.mean = TRUE)
  comparison$AIC[i]        <- round(fit$aic, 1)
  comparison$BIC[i]        <- round(AIC(fit, k = log(length(reg_resid))), 1)
  comparison$LjungBox_p[i] <- round(
    Box.test(residuals(fit), lag = 10, type = "Ljung-Box",
             fitdf = sum(ord[c(1, 3)]))$p.value, 3)
}
comparison
```

## ARIMA(1,0,1) Error Model

$$e_t = \phi_1 e_{t-1} + \varepsilon_t + \theta_1 \varepsilon_{t-1}$$

```{r arima-errors}
sarima(reg_resid, 1, 0, 1)
```

**Model Diagnostics:**

- The standardized residuals show no structure and vary randomly around 0
- The ACF of residuals is within bounds, indicating white noise
- The Q-Q plot appears normal
- The p-values for the Ljung-Box statistic are all above 0.05, indicating no remaining autocorrelation
- Both AR(1) and MA(1) terms are statistically significant (p-values of 0.0002 and 0.0287 respectively)
- AIC = 21.05425, BIC = 21.15595

This is a well-fitting model. Residuals now appear to be white noise, validating the combined regression + ARIMA(1,0,1) approach.

## Combined Model Specification

The final model has a regression component with an intervention at Week 88, and ARIMA(1,0,1) errors to account for autocorrelation in the residuals:

$$y_t = \beta_0 + \beta_1 x_t + \beta_2 t + \beta_3 x_t t + e_t$$

where the error terms $e_t$ follow an ARIMA(1,0,1) process:

$$e_t = \phi_1 e_{t-1} + \varepsilon_t + \theta_1 \varepsilon_{t-1}$$

with:

- $\varepsilon_t$ = white noise
- $\phi_1 = 0.6728$ = autoregressive parameter
- $\theta_1 = -0.4514$ = moving average parameter

This model captures both the structural break at Week 88 and the autocorrelation in the errors. As a robustness check, the regression is re-estimated jointly with the ARIMA(1,0,1) errors by GLS rather than OLS; the post-intervention slope remains strongly negative and of similar magnitude (shown below). The reversal is therefore a real feature of the data rather than an artifact of OLS. The OLS slope is itself unbiased; what the autocorrelation invalidates is its standard errors, not the coefficient.

```{r gls-robustness}
# GLS check: re-fit the regression jointly with the ARIMA(1,0,1) errors and
# compare the post-intervention slope to the OLS estimate.
gls_fit <- arima(
  cereal_ts,
  order = c(1, 0, 1),
  xreg  = cbind(x = cereal$x, t = cereal$t, xt = cereal$xt)
)
ols_slope <- coef(reg_model)["t"] + coef(reg_model)["xt"]
gls_slope <- coef(gls_fit)["t"]  + coef(gls_fit)["xt"]
cat(sprintf("OLS post-intervention slope: %.1f units/week\n", ols_slope))
cat(sprintf("GLS post-intervention slope: %.1f units/week\n", gls_slope))
```

# Forecasting

```{r forecast}
xreg <- cbind(cereal$x, cereal$t, cereal$xt)

h <- 10
future_t  <- (n + 1):(n + h)
future_x  <- ifelse(future_t < 88, 0, 1)
future_xt <- future_x * future_t
newxreg <- cbind(future_x, future_t, future_xt)

fc <- sarima.for(
  x       = cereal_ts,
  n.ahead = h,
  p       = 1,
  d       = 0,
  q       = 1,
  xreg    = xreg,
  newxreg = newxreg,
  plot    = FALSE
)

forecast_df <- data.frame(
  Week     = future_t,
  Forecast = round(fc$pred, 2),
  Lower95  = round(fc$pred - 1.96 * fc$se, 2),
  Upper95  = round(fc$pred + 1.96 * fc$se, 2)
)
forecast_df
```

```{r plot-forecast}
# Base plot: historical data
plot(
  x = 1:length(cereal_ts),
  y = as.numeric(cereal_ts),
  type = "l",
  xlim = c(1, n + h),
  ylim = range(
    c(
      as.numeric(cereal_ts),
      fc$pred + 1.96 * fc$se,
      fc$pred - 1.96 * fc$se
    )
  ),
  main = "Cereal Sales: Historical Data and 10-Week Forecast",
  xlab = "Week",
  ylab = "Sales"
)

# Future x-values (weeks)
future_index <- (n + 1):(n + h)

# Forecast line
lines(future_index, fc$pred, col = "blue", lwd = 2)

# 95% prediction intervals
lines(future_index, fc$pred + 1.96 * fc$se, col = "blue", lty = 2)
lines(future_index, fc$pred - 1.96 * fc$se, col = "blue", lty = 2)

legend(
  "topleft",
  legend = c("Observed", "Forecast", "95% PI"),
  col    = c("black", "blue", "blue"),
  lty    = c(1, 1, 2),
  bty    = "n"
)
```

# Conclusion

This analysis successfully modeled weekly cereal sales with a structural break at week 88 using a combined regression and ARIMA approach.

**Key Findings:**

-   Regression analysis revealed a significant structural break at week 88, coinciding with the competitor's
market entry. The effect is almost entirely a change in trend, not an immediate jump: the level shift at week
88 is small (about -8,100 units, with the raw series flat across the break), while the weekly trend reverses
from +1,153 units/week (growth) to -2,104 units/week (decline), a change in slope of -3,257 units/week
consistent with sustained competitive pressure.
-   Standard regression residuals exhibited significant positive autocorrelation, violating the independence assumption.
-   An ARIMA(1,0,1) error structure effectively captured the autocorrelation, with all diagnostic tests confirming the residuals now behave as white noise.
-   The combined model fits the data well and produces stable 10-week forecasts with reasonable prediction intervals.
