---
title: "Structural Break and Time Series Modeling of Weekly Cereal Sales"
author: "Alex Spigler"
date: "September 2026"
output:
  pdf_document:
    toc: false
    number_sections: false
fontsize: 10pt
geometry: margin=0.75in
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 3.6,
  out.width = "95%"
)

library(car)
library(lmtest)
library(tseries)

break_week <- 88L
season_period <- 52L
forecast_horizon <- 10L

format_estimate <- function(x) {
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

format_p_value <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))
}

make_xreg <- function(week) {
  cbind(
    time = week,
    post_entry = as.integer(week >= break_week),
    weeks_after_entry = pmax(0, week - break_week)
  )
}

linear_combination <- function(fit, terms) {
  estimates <- coef(fit)
  weights <- setNames(numeric(length(estimates)), names(estimates))
  weights[names(terms)] <- terms
  estimate <- sum(weights * estimates)
  standard_error <- sqrt(drop(t(weights) %*% fit$var.coef %*% weights))
  z_value <- estimate / standard_error

  c(
    Estimate = estimate,
    Lower95 = estimate - qnorm(0.975) * standard_error,
    Upper95 = estimate + qnorm(0.975) * standard_error,
    p_value = 2 * pnorm(abs(z_value), lower.tail = FALSE)
  )
}
```

# Introduction

This report examines a published example containing 104 weeks of cereal sales. In the scenario accompanying the data, a competing product enters the market in week 88. The analysis asks whether the sales level or weekly trend changed at that point while accounting for serial correlation.

The dataset comes from the weekly cereal-sales intervention example in Montgomery, Jennings, and Kulahci's *Introduction to Time Series Analysis and Forecasting*, Second Edition, and is available through Wiley's companion materials. The source does not identify the brands or establish whether the values are observed or simulated. This project was originally developed as coursework.

The analysis has three goals:

1. Estimate the immediate level change and the change in trend at week 88.
2. Model autocorrelation rather than relying on ordinary least-squares inference.
3. Produce short-term model-based forecasts and describe their limitations.

# Data and Exploratory Analysis

```{r load-data}
cereal <- read.csv("cereal-sales-data.csv", stringsAsFactors = FALSE)

expected_columns <- c("week", "sales")
stopifnot(
  identical(names(cereal), expected_columns),
  nrow(cereal) == 104L,
  all(is.finite(cereal$week)),
  all(is.finite(cereal$sales)),
  identical(as.integer(cereal$week), seq_len(nrow(cereal))),
  break_week %in% cereal$week
)

cereal$post_entry <- as.integer(cereal$week >= break_week)
cereal$weeks_after_entry <- pmax(0, cereal$week - break_week)
cereal_ts <- ts(cereal$sales, start = 1, frequency = season_period)
xreg <- make_xreg(cereal$week)
```

```{r series-plot, fig.height=3.0}
plot(
  cereal$week,
  cereal$sales,
  type = "l",
  lwd = 1.2,
  main = "Weekly Cereal Sales",
  xlab = "Week",
  ylab = "Sales"
)
abline(v = break_week, col = "firebrick", lty = 2, lwd = 1.5)
text(break_week - 2, min(cereal$sales), "Competitor entry", srt = 90,
     pos = 3, col = "firebrick", cex = 0.8)
```

Sales rise through most of the series and then decline after week 88. A segmented trend is therefore a natural description of the visible change. The event timing is supplied with the example, so the analysis treats week 88 as known rather than searching the series for a break point.

The design is an interrupted time series without a control series or additional covariates. It can estimate a change associated with the stated event date, but it cannot by itself establish that competitor entry caused the change.

# Segmented Regression

Let $I_t$ equal one beginning in week 88, and let $A_t = \max(0, t - 88)$. The deterministic part of the model is

$$
y_t = \beta_0 + \beta_1 t + \beta_2 I_t + \beta_3 A_t + e_t.
$$

This centered parameterization makes the quantities of interest explicit:

- $\beta_1$ is the pre-entry weekly trend.
- $\beta_2$ is the immediate level shift at week 88 relative to the pre-entry trend.
- $\beta_3$ is the change in weekly trend.
- $\beta_1 + \beta_3$ is the post-entry weekly trend.

An ordinary least-squares fit is useful for inspecting the deterministic structure, but its standard errors are not used for final inference if the residuals are autocorrelated.

```{r ols-baseline}
ols_fit <- lm(
  sales ~ time + post_entry + weeks_after_entry,
  data = transform(cereal, time = week)
)
ols_residuals <- residuals(ols_fit)

reset_result <- resettest(ols_fit)
dw_result <- dwtest(ols_fit, alternative = "greater")
ncv_result <- ncvTest(ols_fit)
shapiro_result <- shapiro.test(ols_residuals)
kpss_result <- suppressWarnings(kpss.test(ols_residuals, null = "Level"))

baseline_tests <- data.frame(
  Diagnostic = c(
    "Ramsey RESET",
    "Durbin-Watson (positive autocorrelation)",
    "Nonconstant variance score test",
    "Shapiro-Wilk"
  ),
  p_value = c(
    reset_result$p.value,
    dw_result$p.value,
    ncv_result$p,
    shapiro_result$p.value
  )
)
baseline_tests$p_value <- format_p_value(baseline_tests$p_value)
knitr::kable(baseline_tests, col.names = c("Diagnostic", "p-value"), align = c("l", "r"))
```

```{r baseline-diagnostics, fig.height=4.2}
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(ols_residuals, type = "l", main = "OLS Residuals vs. Time",
     xlab = "Week", ylab = "Residual")
abline(h = 0, col = "firebrick", lty = 2)
plot(fitted(ols_fit), ols_residuals, main = "OLS Residuals vs. Fitted",
     xlab = "Fitted sales", ylab = "Residual")
abline(h = 0, col = "firebrick", lty = 2)
acf(ols_residuals, lag.max = 40, main = "ACF of OLS Residuals")
pacf(ols_residuals, lag.max = 40, main = "PACF of OLS Residuals")
par(old_par)
```

The Durbin-Watson result and the lag-one autocorrelation show that independence is not a reasonable OLS assumption. The other tests and plots do not flag strong nonlinearity, nonconstant variance, or nonnormality, but those p-values are exploratory because they were computed from dependent residuals. The KPSS test does not reject level stationarity ($p > 0.1$), which supports using a stationary error process without differencing; it does not prove stationarity.

Only two annual cycles are observed. The displayed ACF and PACF cover lags through 40, so annual seasonality at lag 52 cannot be assessed reliably from this sample. No seasonal terms are included.

# Joint Regression with ARMA Errors

The regression coefficients and error process are next estimated jointly by maximum likelihood. AR(1), MA(1), and ARMA(1,1) errors are compared because the OLS residual ACF and PACF concentrate most of their structure at short lags.

```{r fit-candidate-models}
candidate_orders <- list(
  "AR(1)" = c(1, 0, 0),
  "MA(1)" = c(0, 0, 1),
  "ARMA(1,1)" = c(1, 0, 1)
)

candidate_fits <- lapply(candidate_orders, function(model_order) {
  arima(
    cereal_ts,
    order = model_order,
    xreg = xreg,
    include.mean = TRUE,
    method = "ML"
  )
})

model_comparison <- data.frame(
  Model = names(candidate_fits),
  AIC = vapply(candidate_fits, AIC, numeric(1)),
  BIC = vapply(candidate_fits, BIC, numeric(1)),
  Ljung_Box_p = vapply(seq_along(candidate_fits), function(i) {
    model_order <- candidate_orders[[i]]
    Box.test(
      residuals(candidate_fits[[i]]),
      lag = 10,
      type = "Ljung-Box",
      fitdf = model_order[1] + model_order[3]
    )$p.value
  }, numeric(1))
)

display_comparison <- transform(
  model_comparison,
  AIC = round(AIC, 1),
  BIC = round(BIC, 1),
  Ljung_Box_p = format_p_value(Ljung_Box_p)
)
knitr::kable(
  display_comparison,
  col.names = c("Error model", "AIC", "BIC", "Ljung-Box p-value (lag 10)"),
  align = c("l", "r", "r", "r")
)
```

ARMA(1,1) has the smallest AIC, but it improves on AR(1) by less than one AIC point; BIC instead favors the simpler AR(1). The data therefore do not identify a uniquely best low-order error model. ARMA(1,1) is retained as the primary specification because it has the lowest AIC and both error-process coefficients are distinguishable from zero. The AR(1) fit is retained as a sensitivity check.

For the selected specification,

$$
e_t = \phi e_{t-1} + \varepsilon_t + \theta \varepsilon_{t-1},
\qquad \varepsilon_t \sim \mathrm{WN}(0, \sigma^2).
$$

```{r selected-model-results}
selected_fit <- candidate_fits[["ARMA(1,1)"]]
ar1_fit <- candidate_fits[["AR(1)"]]

selected_effects <- rbind(
  "Pre-entry weekly trend" = linear_combination(selected_fit, c(time = 1)),
  "Immediate level shift at week 88" = linear_combination(selected_fit, c(post_entry = 1)),
  "Change in weekly trend" = linear_combination(selected_fit, c(weeks_after_entry = 1)),
  "Post-entry weekly trend" = linear_combination(
    selected_fit,
    c(time = 1, weeks_after_entry = 1)
  )
)

effects_table <- data.frame(
  Quantity = rownames(selected_effects),
  Estimate = format_estimate(selected_effects[, "Estimate"]),
  Lower95 = format_estimate(selected_effects[, "Lower95"]),
  Upper95 = format_estimate(selected_effects[, "Upper95"]),
  p_value = format_p_value(selected_effects[, "p_value"]),
  row.names = NULL
)
knitr::kable(
  effects_table,
  col.names = c("Quantity", "Estimate", "95% lower", "95% upper", "p-value"),
  align = c("l", "r", "r", "r", "r")
)

trend_only_fit <- arima(
  cereal_ts,
  order = c(1, 0, 1),
  xreg = cbind(time = cereal$week),
  include.mean = TRUE,
  method = "ML"
)
break_lr_statistic <- 2 * (as.numeric(logLik(selected_fit)) - as.numeric(logLik(trend_only_fit)))
break_lr_p <- pchisq(break_lr_statistic, df = 2, lower.tail = FALSE)

selected_standard_errors <- sqrt(diag(selected_fit$var.coef))
selected_p_values <- 2 * pnorm(
  abs(coef(selected_fit) / selected_standard_errors),
  lower.tail = FALSE
)

ar1_effects <- rbind(
  "Immediate level shift at week 88" = linear_combination(ar1_fit, c(post_entry = 1)),
  "Post-entry weekly trend" = linear_combination(
    ar1_fit,
    c(time = 1, weeks_after_entry = 1)
  )
)

ar_root_modulus <- 1 / abs(coef(selected_fit)["ar1"])
ma_root_modulus <- 1 / abs(coef(selected_fit)["ma1"])
stopifnot(ar_root_modulus > 1, ma_root_modulus > 1)
```

The fitted pre-entry trend is approximately `r format_estimate(selected_effects["Pre-entry weekly trend", "Estimate"])` units per week. The estimated post-entry trend is approximately `r format_estimate(selected_effects["Post-entry weekly trend", "Estimate"])` units per week, a change of `r format_estimate(selected_effects["Change in weekly trend", "Estimate"])` units per week. A likelihood-ratio comparison against an ARMA(1,1) trend-only model supports adding the level and slope changes (p `r format_p_value(break_lr_p)`). The fitted error coefficients are `r sprintf("%.3f", coef(selected_fit)["ar1"])` for AR(1) (p `r format_p_value(selected_p_values["ar1"])`) and `r sprintf("%.3f", coef(selected_fit)["ma1"])` for MA(1) (p = `r format_p_value(selected_p_values["ma1"])`).

The immediate level-shift interval includes zero (p = `r format_p_value(selected_effects["Immediate level shift at week 88", "p_value"])`), so the data do not show a statistically clear jump in week 88. The evidence is concentrated in the change in trend. The AR(1) sensitivity fit produces the same substantive conclusion: an estimated post-entry trend of `r format_estimate(ar1_effects["Post-entry weekly trend", "Estimate"])` units per week and no statistically clear immediate shift (p = `r format_p_value(ar1_effects["Immediate level shift at week 88", "p_value"])`).

```{r selected-model-diagnostics, fig.height=4.6}
selected_residuals <- residuals(selected_fit)
standardized_residuals <- selected_residuals / sqrt(selected_fit$sigma2)
diagnostic_lags <- 3:20
ljung_box_p_values <- vapply(diagnostic_lags, function(model_lag) {
  Box.test(
    selected_residuals,
    lag = model_lag,
    type = "Ljung-Box",
    fitdf = 2
  )$p.value
}, numeric(1))

old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(cereal$week, as.numeric(standardized_residuals), type = "l",
     main = "Standardized Residuals",
     xlab = "Week", ylab = "Residual")
abline(h = 0, col = "firebrick", lty = 2)
acf(as.numeric(standardized_residuals), lag.max = 20, main = "Residual ACF")
qqnorm(as.numeric(standardized_residuals), main = "Normal Q-Q Plot")
qqline(standardized_residuals, col = "firebrick")
plot(diagnostic_lags, ljung_box_p_values, type = "b", ylim = c(0, 1),
     main = "Ljung-Box Diagnostics", xlab = "Lag", ylab = "p-value")
abline(h = 0.05, col = "firebrick", lty = 2)
par(old_par)
```

The selected-model residual ACF has no prominent remaining structure, and the lag-10 Ljung-Box test does not reject residual independence (p = `r format_p_value(model_comparison$Ljung_Box_p[model_comparison$Model == "ARMA(1,1)"])`). These diagnostics are consistent with, but do not prove, a white-noise error sequence. The fitted AR and MA roots lie outside the unit circle, satisfying the stationarity and invertibility conditions.

# Forecasting

The fitted model is projected ten weeks beyond the observed series. These are conditional, model-based forecasts: they assume the segmented post-entry trend and ARMA error structure continue unchanged.

```{r forecasts}
future_weeks <- (nrow(cereal) + 1L):(nrow(cereal) + forecast_horizon)
future_xreg <- make_xreg(future_weeks)
forecast_result <- predict(
  selected_fit,
  n.ahead = forecast_horizon,
  newxreg = future_xreg
)

forecast_table <- data.frame(
  Week = future_weeks,
  Forecast = forecast_result$pred,
  Lower95 = forecast_result$pred - qnorm(0.975) * forecast_result$se,
  Upper95 = forecast_result$pred + qnorm(0.975) * forecast_result$se
)

display_forecasts <- transform(
  forecast_table,
  Forecast = format_estimate(Forecast),
  Lower95 = format_estimate(Lower95),
  Upper95 = format_estimate(Upper95)
)
knitr::kable(
  display_forecasts,
  col.names = c("Week", "Forecast", "95% lower", "95% upper"),
  align = c("r", "r", "r", "r")
)
```

```{r forecast-plot, fig.height=3.0}
plot(
  cereal$week,
  cereal$sales,
  type = "l",
  lwd = 1.2,
  xlim = c(1, max(future_weeks)),
  ylim = range(c(cereal$sales, forecast_table$Lower95, forecast_table$Upper95)),
  main = "Observed Sales and 10-Week Model Projection",
  xlab = "Week",
  ylab = "Sales"
)
polygon(
  c(future_weeks, rev(future_weeks)),
  c(forecast_table$Lower95, rev(forecast_table$Upper95)),
  col = adjustcolor("steelblue", alpha.f = 0.2),
  border = NA
)
lines(future_weeks, forecast_table$Forecast, col = "steelblue4", lwd = 2)
abline(v = break_week, col = "firebrick", lty = 2)
legend(
  "topleft",
  legend = c("Observed", "Forecast", "95% model interval", "Week 88"),
  col = c("black", "steelblue4", adjustcolor("steelblue", alpha.f = 0.3), "firebrick"),
  lty = c(1, 1, 1, 2),
  lwd = c(1.2, 2, 6, 1),
  bty = "n",
  cex = 0.8
)
```

As a limited forecast check, the same ARMA(1,1) specification was refit at ten rolling origins to predict weeks 95 through 104 one step ahead.

```{r rolling-evaluation}
evaluation_weeks <- 95:104
rolling_results <- do.call(rbind, lapply(evaluation_weeks, function(target_week) {
  training_rows <- cereal$week < target_week
  training_fit <- arima(
    ts(cereal$sales[training_rows], start = 1, frequency = season_period),
    order = c(1, 0, 1),
    xreg = make_xreg(cereal$week[training_rows]),
    include.mean = TRUE,
    method = "ML"
  )
  one_step <- predict(
    training_fit,
    n.ahead = 1,
    newxreg = make_xreg(target_week)
  )

  data.frame(
    Week = target_week,
    Actual = cereal$sales[cereal$week == target_week],
    Model = as.numeric(one_step$pred),
    Lower95 = as.numeric(one_step$pred - qnorm(0.975) * one_step$se),
    Upper95 = as.numeric(one_step$pred + qnorm(0.975) * one_step$se),
    Naive = cereal$sales[cereal$week == target_week - 1L]
  )
}))

model_rmse <- sqrt(mean((rolling_results$Actual - rolling_results$Model)^2))
naive_rmse <- sqrt(mean((rolling_results$Actual - rolling_results$Naive)^2))
interval_coverage <- mean(
  rolling_results$Actual >= rolling_results$Lower95 &
    rolling_results$Actual <= rolling_results$Upper95
)

evaluation_table <- data.frame(
  Measure = c("ARMA regression RMSE", "Naive RMSE", "95% interval coverage"),
  Result = c(
    format_estimate(model_rmse),
    format_estimate(naive_rmse),
    sprintf("%d of %d forecasts", sum(
      rolling_results$Actual >= rolling_results$Lower95 &
        rolling_results$Actual <= rolling_results$Upper95
    ), nrow(rolling_results))
  )
)
knitr::kable(evaluation_table, align = c("l", "r"))
```

The model's one-step RMSE is lower than the naive benchmark in this short window, but only `r nrow(rolling_results)` forecasts are evaluated and only `r break_week - 1` pre-entry plus `r min(evaluation_weeks) - break_week` post-entry observations are available at the first origin. Coverage is `r round(100 * interval_coverage)`%, below the nominal 95%. The ten-week projection should therefore be read as an illustration of the fitted model rather than a validated long-range forecast.

# Conclusion

Jointly estimating the segmented regression and ARMA(1,1) errors yields a clear change in weekly trend after week 88. Sales rise by approximately `r format_estimate(selected_effects["Pre-entry weekly trend", "Estimate"])` units per week before the stated entry date and decline by approximately `r format_estimate(abs(selected_effects["Post-entry weekly trend", "Estimate"]))` units per week afterward. The immediate level-shift estimate is uncertain and does not distinguish a jump from zero.

The result is robust to replacing ARMA(1,1) errors with the more parsimonious AR(1) alternative. Residual diagnostics do not flag remaining short-lag autocorrelation, while the limited forecast check shows both modest one-step predictive value and substantial uncertainty. Because this is an uncontrolled textbook example with a short post-entry period, the analysis supports an association with the event timing, not a causal or market-share claim.

# Data Source

Montgomery, D. C., Jennings, C. L., and Kulahci, M. (2015). *Introduction to Time Series Analysis and Forecasting* (2nd ed.). Wiley. Weekly cereal-sales intervention example and [companion data](https://bcs.wiley.com/he-bcs/Books?action=index&bcsId=10036&itemId=1118745116).
