---
title: "Structural Break and Time Series Modeling of Weekly Cereal Sales"
author: "Alex Spigler"
date: "September 2026"
output:
  pdf_document:
    toc: false
    number_sections: false
fontsize: 10pt
geometry: margin=0.75in
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 3.6,
  out.width = "95%"
)

library(car)
library(lmtest)
library(tseries)

break_week <- 88L
forecast_horizon <- 10L

format_estimate <- function(x) {
  format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
}

format_math_estimate <- function(x) {
  gsub(",", "{,}", format_estimate(x), fixed = TRUE)
}

format_math_term <- function(coefficient, symbol) {
  paste(
    if (coefficient < 0) "-" else "+",
    format_math_estimate(abs(coefficient)),
    symbol
  )
}

format_p_value <- function(x) {
  ifelse(x < 0.001, "<0.001", sprintf("%.3f", x))
}

```

# Introduction

This report examines a published example containing 104 weeks of cereal sales. In the scenario accompanying the data, a competing product enters the market in week 88. The analysis asks whether the sales level or weekly trend changed at that point while accounting for serial correlation.

The dataset comes from the weekly cereal-sales intervention example in Montgomery, Jennings, and Kulahci's *Introduction to Time Series Analysis and Forecasting*, Second Edition, and is available through Wiley's companion materials. The source does not identify the brands or establish whether the values are observed or simulated.

The analysis has three goals:

1. Estimate the immediate level change and the change in trend at week 88.
2. Diagnose and model autocorrelation in the regression residuals.
3. Produce short-term model-based forecasts and describe their limitations.

# Data and Exploratory Analysis

```{r load-data}
cereal <- read.csv("cereal-sales-data.csv", stringsAsFactors = FALSE)

expected_columns <- c("week", "sales")
stopifnot(
  identical(names(cereal), expected_columns),
  nrow(cereal) == 104L,
  all(is.finite(cereal$week)),
  all(is.finite(cereal$sales)),
  identical(as.integer(cereal$week), seq_len(nrow(cereal))),
  break_week %in% cereal$week
)

cereal$time <- cereal$week
cereal$post_entry <- as.integer(cereal$week >= break_week)
cereal$weeks_after_entry <- pmax(0, cereal$week - break_week)
```

```{r series-plot, fig.height=3.0}
plot(
  cereal$week,
  cereal$sales,
  type = "l",
  lwd = 1.2,
  main = "Weekly Cereal Sales",
  xlab = "Week",
  ylab = "Sales"
)
abline(v = break_week, col = "firebrick", lty = 2, lwd = 1.5)
```

Sales rise through most of the series and then decline after week 88. A segmented trend is therefore a natural description of the visible change. The event timing is supplied with the example, so the analysis treats week 88 as known rather than searching the series for a break point.

The design is an interrupted time series without a control series or additional covariates. It estimates a change in the brand's weekly sales associated with the stated event date, but it cannot establish that competitor entry caused the change; market share is not observed.

# Segmented Regression

Let $I_t$ equal one beginning in week 88, and let $A_t = \max(0, t - 88)$. The deterministic part of the model is

$$
y_t = \beta_0 + \beta_1 t + \beta_2 I_t + \beta_3 A_t + e_t.
$$

This centered parameterization makes the quantities of interest explicit:

- $\beta_1$ is the pre-entry weekly trend.
- $\beta_2$ is the immediate level shift at week 88 relative to the pre-entry trend.
- $\beta_3$ is the change in weekly trend.
- $\beta_1 + \beta_3$ is the post-entry weekly trend.

The analysis uses a two-step workflow. The segmented regression estimates the level and trend changes, and an ARMA(1,1) model captures the remaining serial dependence in the regression residuals.

```{r ols-baseline}
ols_fit <- lm(
  sales ~ time + post_entry + weeks_after_entry,
  data = cereal
)
ols_residuals <- residuals(ols_fit)
ols_coefficients <- coef(ols_fit)

ols_effects <- c(
  "Pre-entry weekly trend" = unname(ols_coefficients["time"]),
  "Immediate level shift at week 88" = unname(ols_coefficients["post_entry"]),
  "Change in weekly trend" = unname(ols_coefficients["weeks_after_entry"]),
  "Post-entry weekly trend" =
    unname(ols_coefficients["time"] + ols_coefficients["weeks_after_entry"])
)

effects_table <- data.frame(
  Quantity = names(ols_effects),
  Estimate = vapply(ols_effects, format_estimate, character(1)),
  row.names = NULL
)
knitr::kable(
  effects_table,
  col.names = c("Regression quantity", "Estimate"),
  align = c("l", "r"),
  row.names = FALSE
)

reset_result <- resettest(ols_fit)
dw_result <- dwtest(ols_fit, alternative = "greater")
ncv_result <- ncvTest(ols_fit)
shapiro_result <- shapiro.test(ols_residuals)
kpss_result <- suppressWarnings(kpss.test(ols_residuals, null = "Level"))

baseline_tests <- data.frame(
  Diagnostic = c(
    "Ramsey RESET",
    "Durbin-Watson (positive autocorrelation)",
    "Nonconstant variance score test",
    "Shapiro-Wilk"
  ),
  p_value = c(
    reset_result$p.value,
    dw_result$p.value,
    ncv_result$p,
    shapiro_result$p.value
  )
)
baseline_tests$p_value <- format_p_value(baseline_tests$p_value)
knitr::kable(baseline_tests, col.names = c("Diagnostic", "p-value"), align = c("l", "r"))
```

```{r baseline-diagnostics, fig.height=4.2}
old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(ols_residuals, type = "l", main = "OLS Residuals vs. Time",
     xlab = "Week", ylab = "Residual")
abline(h = 0, col = "firebrick", lty = 2)
plot(fitted(ols_fit), ols_residuals, main = "OLS Residuals vs. Fitted",
     xlab = "Fitted sales", ylab = "Residual")
abline(h = 0, col = "firebrick", lty = 2)
acf(ols_residuals, lag.max = 40, main = "ACF of OLS Residuals")
pacf(ols_residuals, lag.max = 40, main = "PACF of OLS Residuals")
par(old_par)
```

The fitted line rises by approximately `r format_estimate(ols_effects["Pre-entry weekly trend"])` units per week before entry and declines by approximately `r format_estimate(abs(ols_effects["Post-entry weekly trend"]))` units per week afterward, a change in slope of `r format_estimate(ols_effects["Change in weekly trend"])` units per week. The immediate level-shift estimate is `r format_estimate(ols_effects["Immediate level shift at week 88"])` units. The primary modeled change is the reversal in trend rather than a one-time shift.

The Durbin-Watson result and the lag-one autocorrelation show that independence is not a reasonable OLS assumption. The other tests and plots do not flag strong nonlinearity, nonconstant variance, or nonnormality, but those p-values are exploratory because they were computed from dependent residuals. The KPSS test found no evidence that the regression residuals are nonstationary ($p > 0.1$), so they are modeled without differencing.

Only two annual cycles are observed. The displayed ACF and PACF cover lags through 40, so annual seasonality at lag 52 cannot be assessed reliably from this sample. No seasonal terms are included.

# Residual Time Series Modeling

The second step models the autocorrelation left in the OLS residuals. AR(1), MA(1), and ARMA(1,1) models are compared because the residual ACF and PACF concentrate most of their structure at short lags. The residual mean is fixed at zero because the regression includes an intercept.

```{r fit-residual-models}
candidate_orders <- list(
  "AR(1)" = c(1, 0, 0),
  "MA(1)" = c(0, 0, 1),
  "ARMA(1,1)" = c(1, 0, 1)
)

candidate_fits <- lapply(candidate_orders, function(model_order) {
  arima(
    ols_residuals,
    order = model_order,
    include.mean = FALSE,
    method = "ML"
  )
})

model_comparison <- data.frame(
  Model = names(candidate_fits),
  AIC = vapply(candidate_fits, AIC, numeric(1)),
  BIC = vapply(candidate_fits, BIC, numeric(1)),
  Ljung_Box_p = vapply(seq_along(candidate_fits), function(i) {
    model_order <- candidate_orders[[i]]
    Box.test(
      residuals(candidate_fits[[i]]),
      lag = 10,
      type = "Ljung-Box",
      fitdf = model_order[1] + model_order[3]
    )$p.value
  }, numeric(1))
)

display_comparison <- transform(
  model_comparison,
  AIC = round(AIC, 1),
  BIC = round(BIC, 1),
  Ljung_Box_p = format_p_value(Ljung_Box_p)
)
knitr::kable(
  display_comparison,
  col.names = c("Residual model", "AIC", "BIC", "Ljung-Box p-value (lag 10)"),
  align = c("l", "r", "r", "r"),
  row.names = FALSE
)
```

ARMA(1,1) has the smallest AIC, while BIC favors the simpler AR(1). Both leave no evident short-lag autocorrelation. ARMA(1,1) is retained for the residual stage because the analysis includes forecasting, both fitted terms contribute within that residual model, and its AIC is lowest. The AIC difference from AR(1) is less than one point, so AR(1) remains a reasonable parsimonious alternative rather than a decisively inferior model.

```{r selected-residual-model}
selected_error_fit <- candidate_fits[["ARMA(1,1)"]]
selected_error_coefficients <- coef(selected_error_fit)
selected_error_standard_errors <- sqrt(diag(selected_error_fit$var.coef))
selected_error_p_values <- 2 * pnorm(
  abs(selected_error_coefficients / selected_error_standard_errors),
  lower.tail = FALSE
)

error_parameter_table <- data.frame(
  Parameter = c("AR(1)", "MA(1)"),
  Estimate = sprintf("%.3f", selected_error_coefficients[c("ar1", "ma1")]),
  Standard_error = sprintf("%.3f", selected_error_standard_errors[c("ar1", "ma1")]),
  p_value = format_p_value(selected_error_p_values[c("ar1", "ma1")])
)
knitr::kable(
  error_parameter_table,
  col.names = c("Residual parameter", "Estimate", "Standard error", "p-value"),
  align = c("l", "r", "r", "r"),
  row.names = FALSE
)

ar_root_modulus <- 1 / abs(selected_error_coefficients["ar1"])
ma_root_modulus <- 1 / abs(selected_error_coefficients["ma1"])
stopifnot(ar_root_modulus > 1, ma_root_modulus > 1)
```

The two fitted stages are

$$
\begin{aligned}
\widehat{y}^{\mathrm{reg}}_t ={}& `r format_math_estimate(ols_coefficients["(Intercept)"])`
`r format_math_term(ols_coefficients["time"], "t")`
`r format_math_term(ols_coefficients["post_entry"], "I_t")`
`r format_math_term(ols_coefficients["weeks_after_entry"], "A_t")`, \\
r_t ={}& `r sprintf("%.3f", selected_error_coefficients["ar1"])`r_{t-1}
`r if (selected_error_coefficients["ma1"] < 0) "-" else "+"`
`r sprintf("%.3f", abs(selected_error_coefficients["ma1"]))`\varepsilon_{t-1}
+ \varepsilon_t,
\qquad \varepsilon_t \sim \mathrm{WN}(0, \widehat{\sigma}^2).
\end{aligned}
$$

Here $r_t$ denotes the residual from the segmented regression. The ARMA(1,1) model captures the serial dependence in those residuals. Diagnostics of the resulting ARMA residuals show no evidence of remaining autocorrelation, strong nonnormality, or conditional heteroskedasticity.

```{r residual-model-tests}
arma_residuals <- residuals(selected_error_fit)
standardized_arma_residuals <- as.numeric(
  arma_residuals / sqrt(selected_error_fit$sigma2)
)

selected_ljung_box_result <- Box.test(
  arma_residuals,
  lag = 10,
  type = "Ljung-Box",
  fitdf = 2
)
selected_shapiro_result <- shapiro.test(standardized_arma_residuals)

arch_lags <- 5L
arch_matrix <- embed(standardized_arma_residuals^2, arch_lags + 1L)
arch_response <- arch_matrix[, 1]
arch_predictors <- arch_matrix[, -1, drop = FALSE]
arch_fit <- lm(arch_response ~ arch_predictors)
arch_statistic <- nrow(arch_matrix) * summary(arch_fit)$r.squared
arch_p_value <- pchisq(arch_statistic, df = arch_lags, lower.tail = FALSE)

selected_diagnostic_tests <- data.frame(
  Diagnostic = c(
    "Ljung-Box ARMA residual autocorrelation (lag 10)",
    "Shapiro-Wilk normality",
    "ARCH LM conditional heteroskedasticity (5 lags)"
  ),
  p_value = format_p_value(c(
    selected_ljung_box_result$p.value,
    selected_shapiro_result$p.value,
    arch_p_value
  ))
)
knitr::kable(
  selected_diagnostic_tests,
  col.names = c("Residual-model diagnostic", "p-value"),
  align = c("l", "r"),
  row.names = FALSE
)
```

```{r residual-model-diagnostics, fig.height=4.6}
diagnostic_lags <- 3:20
ljung_box_p_values <- vapply(diagnostic_lags, function(model_lag) {
  Box.test(
    arma_residuals,
    lag = model_lag,
    type = "Ljung-Box",
    fitdf = 2
  )$p.value
}, numeric(1))

old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(cereal$week, standardized_arma_residuals, type = "l",
     main = "Standardized ARMA Residuals",
     xlab = "Week", ylab = "Residual")
abline(h = 0, col = "firebrick", lty = 2)
acf(standardized_arma_residuals, lag.max = 20, main = "ARMA Residual ACF")
qqnorm(standardized_arma_residuals, main = "Normal Q-Q Plot")
qqline(standardized_arma_residuals, col = "firebrick")
plot(diagnostic_lags, ljung_box_p_values, type = "b", ylim = c(0, 1),
     main = "Ljung-Box Diagnostics", xlab = "Lag", ylab = "p-value")
abline(h = 0.05, col = "firebrick", lty = 2)
par(old_par)
```

The lag-10 Ljung-Box test finds no evidence of remaining autocorrelation through lag 10 (p = `r format_p_value(selected_ljung_box_result$p.value)`). The Shapiro-Wilk test and Q-Q plot do not flag strong nonnormality (p = `r format_p_value(selected_shapiro_result$p.value)`), and the five-lag ARCH test does not flag conditional heteroskedasticity (p = `r format_p_value(arch_p_value)`). The fitted AR and MA roots satisfy the stationarity and invertibility conditions.

# Forecasting

The point forecast adds the segmented-regression projection to the forecast of its ARMA(1,1) residual. These projections assume that the post-entry trend and residual process continue unchanged. The intervals reflect uncertainty from the residual process only; they treat the regression coefficients and fitted ARMA parameters as fixed and exclude model-selection uncertainty.

```{r forecasts}
future_weeks <- (nrow(cereal) + 1L):(nrow(cereal) + forecast_horizon)
future_data <- data.frame(
  time = future_weeks,
  post_entry = as.integer(future_weeks >= break_week),
  weeks_after_entry = pmax(0, future_weeks - break_week)
)
deterministic_forecast <- as.numeric(predict(ols_fit, newdata = future_data))
residual_forecast <- predict(selected_error_fit, n.ahead = forecast_horizon)
combined_forecast <- deterministic_forecast + as.numeric(residual_forecast$pred)

forecast_table <- data.frame(
  Week = future_weeks,
  Forecast = combined_forecast,
  Lower95 = combined_forecast - qnorm(0.975) * residual_forecast$se,
  Upper95 = combined_forecast + qnorm(0.975) * residual_forecast$se
)

display_forecasts <- transform(
  forecast_table,
  Forecast = format_estimate(Forecast),
  Lower95 = format_estimate(Lower95),
  Upper95 = format_estimate(Upper95)
)
knitr::kable(
  display_forecasts,
  col.names = c("Week", "Forecast", "Cond. 95% lower", "Cond. 95% upper"),
  align = c("r", "r", "r", "r")
)
```

```{r forecast-plot, fig.height=3.0}
plot(
  cereal$week,
  cereal$sales,
  type = "l",
  lwd = 1.2,
  xlim = c(1, max(future_weeks)),
  ylim = range(c(cereal$sales, forecast_table$Lower95, forecast_table$Upper95)),
  main = "Observed Sales and 10-Week Conditional Projection",
  xlab = "Week",
  ylab = "Sales"
)
polygon(
  c(future_weeks, rev(future_weeks)),
  c(forecast_table$Lower95, rev(forecast_table$Upper95)),
  col = adjustcolor("steelblue", alpha.f = 0.2),
  border = NA
)
lines(future_weeks, forecast_table$Forecast, col = "steelblue4", lwd = 2)
abline(v = break_week, col = "firebrick", lty = 2)
legend(
  "topleft",
  legend = c("Observed", "Forecast", "Conditional 95% interval", "Week 88"),
  col = c("black", "steelblue4", adjustcolor("steelblue", alpha.f = 0.3), "firebrick"),
  lty = c(1, 1, 1, 2),
  lwd = c(1.2, 2, 6, 1),
  bty = "n",
  cex = 0.8
)
```

As a limited forecast check, both stages were refit at ten rolling origins to predict weeks 95 through 104 one step ahead. The segmented regression is identical across the two candidates; only the fitted residual model changes.

```{r rolling-evaluation}
evaluation_weeks <- 95:104

fit_two_step_one_ahead <- function(target_week, residual_order) {
  training_rows <- cereal$week < target_week
  training_data <- cereal[training_rows, ]
  target_data <- cereal[cereal$week == target_week, ]

  training_regression <- lm(
    sales ~ time + post_entry + weeks_after_entry,
    data = training_data
  )
  training_error_fit <- arima(
    residuals(training_regression),
    order = residual_order,
    include.mean = FALSE,
    method = "ML"
  )
  regression_prediction <- as.numeric(
    predict(training_regression, newdata = target_data)
  )
  error_prediction <- predict(training_error_fit, n.ahead = 1)
  combined_prediction <- regression_prediction + as.numeric(error_prediction$pred)

  c(
    Prediction = combined_prediction,
    Lower95 = combined_prediction - qnorm(0.975) * as.numeric(error_prediction$se),
    Upper95 = combined_prediction + qnorm(0.975) * as.numeric(error_prediction$se)
  )
}

rolling_results <- do.call(rbind, lapply(evaluation_weeks, function(target_week) {
  ar1_forecast <- fit_two_step_one_ahead(target_week, c(1, 0, 0))
  arma_forecast <- fit_two_step_one_ahead(target_week, c(1, 0, 1))

  data.frame(
    Week = target_week,
    Actual = cereal$sales[cereal$week == target_week],
    AR1 = ar1_forecast["Prediction"],
    AR1_Lower95 = ar1_forecast["Lower95"],
    AR1_Upper95 = ar1_forecast["Upper95"],
    ARMA = arma_forecast["Prediction"],
    ARMA_Lower95 = arma_forecast["Lower95"],
    ARMA_Upper95 = arma_forecast["Upper95"],
    Naive = cereal$sales[cereal$week == target_week - 1L]
  )
}))

ar1_rmse <- sqrt(mean((rolling_results$Actual - rolling_results$AR1)^2))
arma_rmse <- sqrt(mean((rolling_results$Actual - rolling_results$ARMA)^2))
naive_rmse <- sqrt(mean((rolling_results$Actual - rolling_results$Naive)^2))
ar1_interval_hits <- sum(
  rolling_results$Actual >= rolling_results$AR1_Lower95 &
    rolling_results$Actual <= rolling_results$AR1_Upper95
)
arma_interval_hits <- sum(
  rolling_results$Actual >= rolling_results$ARMA_Lower95 &
    rolling_results$Actual <= rolling_results$ARMA_Upper95
)

evaluation_table <- data.frame(
  Measure = c(
    "Two-step AR(1) RMSE",
    "Two-step ARMA(1,1) RMSE",
    "Naive RMSE",
    "AR(1) residual-only interval coverage",
    "ARMA(1,1) residual-only interval coverage"
  ),
  Result = c(
    format_estimate(ar1_rmse),
    format_estimate(arma_rmse),
    format_estimate(naive_rmse),
    sprintf("%d of %d forecasts", ar1_interval_hits, nrow(rolling_results)),
    sprintf("%d of %d forecasts", arma_interval_hits, nrow(rolling_results))
  )
)
knitr::kable(evaluation_table, align = c("l", "r"))
```

Both two-step forecasts outperform the naive benchmark in this short window. ARMA(1,1) is `r format_estimate(ar1_rmse - arma_rmse)` RMSE units (`r sprintf("%.1f", 100 * (ar1_rmse - arma_rmse) / ar1_rmse)`%) below AR(1), which is too small a difference to distinguish the residual specifications decisively. Each covers `r arma_interval_hits` of 10 residual-only intervals. Because the evaluation contains only `r nrow(rolling_results)` forecasts---with `r break_week - 1` pre-entry and `r min(evaluation_weeks) - break_week` post-entry observations at the first origin---and the intervals omit regression and parameter-estimation uncertainty, the projection is illustrative rather than a validated long-range forecast.

# Conclusion

The segmented regression estimates a pre-entry rise of `r format_estimate(ols_effects["Pre-entry weekly trend"])` units per week and a post-entry decline of `r format_estimate(abs(ols_effects["Post-entry weekly trend"]))` units per week, with a slope change of `r format_estimate(ols_effects["Change in weekly trend"])` units per week. The estimated immediate level shift is `r format_estimate(ols_effects["Immediate level shift at week 88"])` units; the primary modeled change is the trend reversal rather than a one-time jump.

An ARMA(1,1) model removes the evident short-lag autocorrelation from the regression residuals. AR(1) is also adequate, and the limited rolling check does not establish a meaningful forecasting advantage for either residual model. This uncontrolled example with a short post-entry period supports a descriptive association with the stated event date, not a causal conclusion; market share is not observed.

# Data Source

Montgomery, D. C., Jennings, C. L., and Kulahci, M. (2015). *Introduction to Time Series Analysis and Forecasting* (2nd ed.). Wiley. Weekly cereal-sales intervention example and [companion data](https://bcs.wiley.com/he-bcs/Books?action=index&bcsId=10036&itemId=1118745116).
